OR Multi Disturbance, NIMBLE version w/ Wishart

### 2025-11-21 OR_model_DisturbPriorNIMBLE.Rmd
in-progress versions of OR model that adds disturbance product priors

---
title: "OR Multi Disturbance"
author: "Michael Dietze"
date: "2024-05-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("disturbance.R")
source("multi.ens.adj.R")
library(BayesianTools)
```

@@

## functions
```{r,echo=FALSE,results='hide',message=FALSE}
##' Single Site, Single Disturbance
##' @param Y observed data
##' @param R observation error
##' @param Q process error
##' @param Xic state initial condition matrix
##' @param THETA parameter matrix
##' @param dist  disturbance list: p.dist[1:NT], mu0[3], V0[3,3], alloc soil 
SSSD.model <- function(Y,R,Q,PAR,Xic,THETA,dist){

  ## init
  NT = length(PAR)/365
  Nf = Nmu = Na = Xic
  ne = nrow(Xic)
  if(!is.matrix(THETA)) THETA = matrix(THETA,nrow=ne,ncol=length(THETA),byrow = TRUE)
  Forecast = Analysis = list()

  AnalysisSSSDMP <- "  ## SingleSite SingleDisturbance MultiPool
    model{
      ## prior
      muN ~ dmnorm(mufN,pfN)                ## forecast prior on Not disturbed
      muD ~ dmnorm(mufD,pfD)                ## forecast prior on Disturbed
      D ~ dbern(p)                          ## disturbance probability
      ## Analysis 
      n <- D*muD + (1-D)*muN                ## select between disturbed and undisturbed
      Y[H] ~ dmnorm(n[H],R[H,H])                       ## Observation error
  }"

  for(t in 1:NT){

@@

    ## Forecast 
    for(i in 1:ne){
      tsel    = (t-1)*365 + 1:365             ## Select days from driver files
      Nmu[i,] = VSEM(THETA[i,], PAR = PAR[tsel])[365,2:4] ## deterministic simulation
      Nf[i,]  = mvtnorm::rmvnorm(1,Nmu[i,],Q) ## process error
    }

    ## Disturbance
    d.forecast <- rep(0,ne); d.forecast[1:ceiling(ne*0.1)] <- 1 #rbinom(ne,1,dist$p.dist)             ## prior on disturbance status
    sel <- !as.logical(d.forecast)                ## select undisturbed ensemble members
    if(sum(!sel)>2){
      sel.dist = which(!sel)
      for(j in sel.dist){                      ## run disturbance model
        Nf[j,]  = disturbance.p(Nf[j,],mu0=dist$mu0,V0 = dist$V0,alloc.soil = dist$alloc.soil)
      }
      mufD  = colMeans(Nf[!sel,])
      pfD = solve(var(Nf[!sel,]))#solve(diag(c(0.15,0.5,0.25)^2))
    } 
    if(sum(!sel) <= 2 || "try-error" %in% class(pfD)) {   ## backup values if the ensemble stats fail
      mufD = c(dist$mu0[1], mean(Nf[,2]),  dist$mu0[2])            
      pfD  = solve(diag(c(0.15,0.5,0.25)^2))
    }
    Nf <- tobit(Nf)

    ## priors
    priors <- list(p    = dist$p.dist[t],         ## disturbance rate
                   mufN = colMeans(Nf[sel,]),    ## Not disturbed forecast means
                   mufD = mufD,   ## Disturbed forecast means
                   pfN  = solve(var(Nf[sel,])),  ## Not disturbed forecast precision
                   pfD  = pfD  ## Disturbed forecast precision
    )
    Forecast[[t]] <- c(list(d.forecast=d.forecast,Nf=Nf),priors)

    ## Analysis
    update = c(list(Y=Y[t,],R=solve(R)),priors,H = which(!is.na(Y[t,])))
    if(length(update$H) > 0){ ## data observed, run filter
      mod <- rjags::jags.model(file=textConnection(AnalysisSSSDMP),
                               data=update,
                               n.adapt=1000,n.chains=3,quiet=TRUE)
      jdat <- rjags::coda.samples(mod,variable.names=c("n","D","muN","muD"),n.iter=10000)

      ## update parameters & save Analysis posterior
      dat = as.matrix(jdat)
      nsel <- grep(pattern="^n",x=colnames(dat))
      Xa = dat[,nsel]
      ca = dat[,"D"]
      run.adjustment = TRUE
    } else {
      Xa = Nf
      ca = 1
      run.adjustment = FALSE
    }

      ## draw new IC
    if(run.adjustment){
    Na <- multi.ens.adj(
      Xf   = Nf,
      cf   = d.forecast,
      mu.f = rbind(priors$mufN,priors$mufD),
      Pf   = abind::abind(solve(priors$pfN),
                            solve(priors$pfD),
                            along=3),
      Xa = Xa,
      ca = ca)
    } else {
      Na = cbind(rep(0,nrow(Nf)),Nf)
    }
    Ca <- Na[,1] ## updated classf assignments
    Na <- tobit(Na[,-1])
    #  sel.ic <- sample.int(nrow(dat), ne, replace=TRUE)
    #  Na <- dat[sel.ic,nsel]
    Analysis[[t]] <- list(Na = Na,
                          Ca = Ca,
                          CI = apply(dat,2,quantile,c(0.025,0.25,0.5,0.75,0.975)),
                          mu = apply(dat,2,mean,na.rm=TRUE)
    )
  }
  return(list(Forecast=Forecast,Analysis=Analysis))
}
```

@@

## analyze sites - just one disturbance
```{r}
Rmult = c(1,0.1,0.01,0.001)
Qmult = c(1,10)
Pmult = c(1)
Ri = 2
Qi = 1
Pi = 1
users_accuracy = c(1-d.prior[1],.77,.92,.77,.78,.78) #Kennedy et al 2015 table 6
## 0 = none, 10 = other, 20 = cut, 30 = road, 40 = fire, 50 = biotic
od = c(83:176,264:356)
for(i in 1:356){
  ofile = paste0("OR_DP_",  ## DP = Disturbance Prior
                 formatC(i, width = 3, format = "d", flag = "0"),
                 "_P",Pi,"Q",Qi,"R",Ri,".RData")
  print(ofile)
  if(file.exists(ofile)) next
  NT = floor(ncol(rad)/365)
  Yi = matrix(NA,NT,3)  ## observed data
  Yi[1:ncol(Yw),3] = Yw[i,]  ## only Cw observed
  R = diag(rep(Rw,3))*Rmult[Ri]    ## observation error
  Q = diag(c(Ql,Qsoil,Qw)^2)*Qmult[Qi]  ## process error
  dtype = which(names(d.prior) == dextract[i,"t2004"])
  if(dtype == 1) dtype = sample(2:5,1) ## if site is undisturbed, set alternative at random
  if(length(dtype)==0) break
  p.dist = rep(d.prior[dtype]*Pmult[Pi],NT) ## set background rate
  has.dist = which(dextract[i,] > 0)
  p.dist[has.dist] = users_accuracy[dextract[i,has.dist]/10 + 1] 
  dist <- list(              ## disturbance parameters
    #p.dist = d.prior[dtype]*Pmult[Pi],     ## disturbance probability
    p.dist = p.dist,
    mu0 = c(0.2,PBBAR[4,dtype]),   ## % retention [leaf,agb], first parameter made up, should include soil
    V0 =  diag(c(0.15,PBVAR[4,dtype])^2), #variance in retention
    alloc.soil = 0.05        ## made up  
  )
  test = SSSD.model(Y = Yi,R=R,Q=Q,
                  PAR=rad[i,],Xic=ICprior,THETA=theta[1:11],
                  dist = dist)
  save(test,dist,file=ofile)
  test$Analysis[[20]]$mu
}
```

## visualization
```{r}
i = 188#181
ofile = paste0("OR_DP_",  ## DP = Disturbance Prior
                 formatC(i, width = 3, format = "d", flag = "0"),
                 "_P",Pi,"Q",Qi,"R",Ri,".RData")
load(ofile)
NT = floor(ncol(rad)/365)
Yi = matrix(NA,NT,3)  ## observed data
Yi[1:ncol(Yw),3] = Yw[i,]  ## only Cw observed
pool.names=c("Cleaf","Csoil","Cstem")
for(i in 1:3){

@@

  plot(Yi[,i],
       ylim = range(c(range(rbind(CI.Na,Yi[,i]),na.rm=TRUE),0)),
       type='n',
       ylab=pool.names[i],xlab="time")
  ecoforecastR::ciEnvelope(1:ncol(CI.Nf),CI.Nf[1,],CI.Nf[3,],col=ecoforecastR::col.alpha("blue",0.5))
  ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[5,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
  lines(CI.Na[3,],col="#b2df8a",lwd=2)
  points(Yi[,i])
  lines(CI.D[3,])

@@

>>>>

---
title: "OR Multi Disturbance, NIMBLE version w/ Wishart"
author: "Michael Dietze"
date: "2024-05-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("disturbance.R")
source("multi.ens.adj.R")
library(BayesianTools)
library(nimble)
```

@@

## functions
```{r,echo=FALSE,results='hide',message=FALSE}
##' Single Site, Single Disturbance
##' @param Y observed data
##' @param R observation error
##' @param Q process error
##' @param Xic state initial condition matrix
##' @param THETA parameter matrix
##' @param dist  disturbance list: p.dist[1:NT], mu0[3], V0[3,3], alloc soil 
SSSD.model <- function(Y,R,Q,PAR,Xic,THETA,dist,restart=FALSE){

  uni = TRUE

  ## init
  NT = length(PAR)/365
  Nf = Nmu = Na = Xic
  ne = nrow(Xic)
  if(!is.matrix(THETA)) THETA = matrix(THETA,nrow=ne,ncol=length(THETA),byrow = TRUE)
  Forecast = Analysis = list()
  bq.stored <- rep(NA,NT)
  bq.stored[1] <- ifelse(uni,1,4)
  if(uni){
    aq.stored <- rep(NA,NT)
    aq.stored[1] <- Q[3,3]*bq.stored[1]
  } else {
    aq.stored <- array(NA, c(NT,3,3))
    aq.stored[1,,] <- Q*bq.stored[1]
  }
  t0 = 1

  ## restart
  if(restart){
    load("test.RData")
    Na = Analysis[[t]]$Na
    t0 = t
  }

  AnalysisSSSDMP.uni <- nimble::nimbleCode({  ## univariate constraint
    # Prior
    muN[1:N] ~ dmnorm(mufN[1:N],prec=pfN[1:N,1:N]) ## forecast prior on Not disturbed
    muD[1:N] ~ dmnorm(mufD[1:N],prec=pfD[1:N,1:N]) ## forecast prior on Disturbed
    D ~ dbern(prob=p)                         ## disturbance probability
    q ~ dgamma(bq,aq) ## wishart prior precision 
    ## Likelihood
    n[1:N] <- D*muD[1:N] + (1-D)*muN[1:N]     ## select between disturbed and undisturbed
    X ~ dnorm(n[H], 1/sqrt(q))                ## add process error  
    Y ~ dnorm(X,R)                            ## Observation error
  })

    AnalysisSSSDMP <- nimble::nimbleCode({
    # Prior
    muN[1:N] ~ dmnorm(mufN[1:N],prec=pfN[1:N,1:N]) ## forecast prior on Not disturbed
    muD[1:N] ~ dmnorm(mufD[1:N],prec=pfD[1:N,1:N]) ## forecast prior on Disturbed
    D ~ dbern(prob=p)                         ## disturbance probability
    q[1:N, 1:N]  ~ dwish(R = aq[1:N, 1:N], df = bq) ## wishart prior precision
    ## Likelihood
    n[1:N] <- D*muD[1:N] + (1-D)*muN[1:N]     ## select between disturbed and undisturbed
    #X[1:N]  ~ dmnorm(n[1:N], prec = q[1:N, 1:N])  ## add process error
    X  ~ dnorm(n[H], 1/sqrt(q[H, H]))  ## add process error  ## HACK
    #if(NY == 1){
    #  Y ~ dnorm(X[H],R)                        ## Observation error
    Y ~ dnorm(X,R)                        ## Observation error ## HACK
    #} else {
    #  for(i in 1:NY){                            
    #    ns[i] <-X[H[i]]                        ## Observation operator
    #  }
    #  Y[1:NY] ~ dmnorm(ns[1:NY],R[1:NY,1:NY])    ## Observation error
    #}
  })

  for(t in t0:NT){

@@

    ## Forecast 
    for(i in 1:ne){
      tsel    = (t-1)*365 + 1:365             ## Select days from driver files
      Nf[i,] = VSEM(THETA[i,], PAR = PAR[tsel])[365,2:4] ## deterministic simulation
      ##Nf[i,]  = mvtnorm::rmvnorm(1,Nmu[i,],Q) ## process error
      ## Nimble Wishart version handles process error internally
    }

    ## Disturbance
    d.forecast <- rep(0,ne); d.forecast[1:ceiling(ne*0.1)] <- 1 #rbinom(ne,1,dist$p.dist)             ## prior on disturbance status
    sel <- !as.logical(d.forecast)                ## select undisturbed ensemble members
    if(sum(!sel)>2){
      sel.dist = which(!sel)
      for(j in sel.dist){                      ## run disturbance model
        Nf[j,]  = disturbance.p(Nf[j,],mu0=dist$mu0,V0 = dist$V0,alloc.soil = dist$alloc.soil)
      }
      mufD  = colMeans(Nf[!sel,])
      pfD = try(solve(var(Nf[!sel,])))#solve(diag(c(0.15,0.5,0.25)^2))
    } 
    if(sum(!sel) <= 2 || "try-error" %in% class(pfD)) {   ## backup values if the ensemble stats fail
      mufD = c(dist$mu0[1], mean(Nf[,2]),  dist$mu0[2])            
      pfD  = solve(diag(c(0.15,0.5,0.25)^2))
    }
    mufN  =colMeans(Nf[sel,])
    varfN = var(Nf[sel,])
    tooSmall = which(diag(varfN)/mufN < 0.001)
    for(jj in seq_along(tooSmall)){
      varfN[tooSmall[jj],tooSmall[jj]] = mufN[tooSmall[jj]]*0.001
    }
    pfN   = solve(varfN)
    #if("try-error" %in% class(pfD)){
    #  stop()
    #}
    Nf <- tobit(Nf)

    ## priors
    aq = ifelse(uni,aq.stored[t],aq.stored[t,,])
    priors <- list(p    = dist$p.dist[t],       ## disturbance rate
                   mufN = mufN,   ## Not disturbed forecast means
                   mufD = mufD,                 ## Disturbed forecast means
                   pfN  = pfN,                  ## Not disturbed forecast precision
                   pfD  = pfD,                  ## Disturbed forecast precision
                   aq   = aq,                   ## prior sum of squares
                   bq   = bq.stored[t]          ## prior sample size
    )
    Forecast[[t]] <- c(list(d.forecast=d.forecast,Nf=Nf),priors)

    ## Analysis
    update = c(list(Y=Y[t,3],         ## Data                #### HACK
                    R=solve(R)[3,3]), ## Observation error   #### HACK
                    priors)
    constants = list(N=3,                      ## number of state variables
                     H = as.vector(which(!is.na(Y[t,]))))  ## observation vector
    constants$NY = length(constants$H)
    if(constants$NY > 0){ ## data observed, run filter
      mod <- nimble::nimbleModel(code = AnalysisSSSDMP.uni, 
                                 data = update, 
                                 constants=constants)
      conf <- nimble::configureMCMC(mod, print=FALSE)
      conf$addMonitors(c("X","n","D","q"))   #,"muN","muD"
      conf$printSamplers()
      Rmcmc <- nimble::buildMCMC(conf)
      Cmodel <- nimble::compileNimble(mod)
      Cmcmc <- nimble::compileNimble(Rmcmc, project = mod, showCompilerOutput = FALSE)
      dat  <- nimble::runMCMC(Cmcmc, niter = 10000, 
                               nburnin = 3000, thin = 1, 
                               nchains = 3,samplesAsCodaMCMC = TRUE)

      ## update parameters & save Analysis posterior
      dat = as.matrix(dat)
      nsel <- grep(pattern="^n",x=colnames(dat))
      xsel <- grep(pattern="^X",x=colnames(dat))
      if(uni){
        Xa = dat[,nsel]
        Xa[,3] = dat[,xsel]
      } else {
        Xa = dat[,xsel]
      }
      ca = dat[,"D"]
      ## update Wishart
      iq   <- grep("q", colnames(dat)) # 9 =3*3 (mean of those columns to generate a vector of 9)
      mq <- dat[, iq]  # Omega, Precision
      dim.q <- sqrt(length(iq))
      if(uni){
        q.bar = mean(mq) ## currently all 1 very small #
      } else {
        q.bar <- matrix(apply(mq, 2, mean), dim.q, dim.q)  # Mean Omega, Precision
      }
      col <- matrix(1:dim.q ^ 2, dim.q, dim.q)
      WV  <- matrix(0, dim.q, dim.q)
      if(uni){
        WV = q.bar/var(mq)
      }else{
        wish.df <- function(Om, X, i, j, col) { # degrees of freedom function  
          (Om[i, j]^2 + Om[i, i] * Om[j, j]) / stats::var(X[, col])
        }
        for (ii in seq_len(dim.q)) {
          for (jj in seq_len(dim.q)) {
            WV[ii, jj] <- wish.df(q.bar, X = mq, i = ii, j = jj, col = col[ii, jj])
          }
        }
      }
      mean.WV <- mean(WV)
      if (mean.WV < dim.q) {
         mean.WV <- dim.q
      }
      V <- solve(q.bar) * mean.WV
      if(uni){
        aq.stored[t+1]       <- V
      } else {
        aq.stored[t + 1, , ] <- V
      }
      bq.stored[t + 1]       <- mean.WV

      run.adjustment = TRUE
    } else {
      Xa = Nf
      ca = 1
      run.adjustment = FALSE
    }
#hist(1/sqrt(mq))

      ## draw new IC
    if(run.adjustment){
    Na <- multi.ens.adj(
      Xf   = Nf,
      cf   = d.forecast,
      mu.f = rbind(priors$mufN,priors$mufD),
      Pf   = abind::abind(solve(priors$pfN),
                            solve(priors$pfD),
                            along=3),
      Xa = Xa,
      ca = ca)
    } else {
      Na = cbind(rep(0,nrow(Nf)),Nf)
    }
    Ca <- Na[,1] ## updated classf assignments
    Na <- tobit(Na[,-1])
    #  sel.ic <- sample.int(nrow(dat), ne, replace=TRUE)
    #  Na <- dat[sel.ic,nsel]
    Analysis[[t]] <- list(Na = Na,
                          Ca = Ca,
                          CI = apply(dat,2,quantile,c(0.025,0.25,0.5,0.75,0.975)),
                          mu = apply(dat,2,mean,na.rm=TRUE)
    )
    save(Forecast,Analysis,aq.stored, bq.stored,
         t,Y,R,Q,PAR,THETA,dist,file="test.RData")
    # test = list(Forecast=Forecast,Analysis=Analysis,aq=aq.stored,bq=bq.stored)
  }
  return(list(Forecast=Forecast,Analysis=Analysis,aq=aq.stored,bq=bq.stored))
}
```

@@

## analyze sites - just one disturbance
```{r}
Rmult = c(1,0.1,0.01,0.001)
Qmult = c(1,10)
Pmult = c(1)
Ri = 1
Qi = 1
Pi = 1
users_accuracy = c(1-d.prior[1],.77,.92,.77,.78,.78) #Kennedy et al 2015 table 6
## 0 = none, 10 = other, 20 = cut, 30 = road, 40 = fire, 50 = biotic
od = c(83:176,264:356)
for(i in 181:280){
  ofile = paste0("OR_NDP_",  ## N = Nimble, DP = Disturbance Prior
                 formatC(i, width = 3, format = "d", flag = "0"),
                 "_P",Pi,"Q",Qi,"R",Ri,".RData")
  print(ofile)
  if(file.exists(ofile)) next
  NT = floor(ncol(rad)/365)
  Yi = matrix(NA,NT,3)  ## observed data
  Yi[1:ncol(Yw),3] = Yw[i,]  ## only Cw observed
  R = diag(rep(Rw,3))*Rmult[Ri]    ## observation error
  Q = diag(c(Ql,Qsoil,Qw)^2)*Qmult[Qi]  ## process error
  dtype = which(names(d.prior) == dextract[i,"t2004"])
  if(dtype == 1) dtype = sample(2:5,1) ## if site is undisturbed, set alternative at random
  if(length(dtype)==0) break
  p.dist = rep(d.prior[dtype]*Pmult[Pi],NT) ## set background rate
  has.dist = which(dextract[i,] > 0)
  p.dist[has.dist] = users_accuracy[dextract[i,has.dist]/10 + 1] 
  dist <- list(              ## disturbance parameters
    #p.dist = d.prior[dtype]*Pmult[Pi],     ## disturbance probability
    p.dist = p.dist,
    mu0 = c(0.2,PBBAR[4,dtype]),   ## % retention [leaf,agb], first parameter made up, should include soil
    V0 =  diag(c(0.15,PBVAR[4,dtype])^2), #variance in retention
    alloc.soil = 0.05        ## made up  
  )
  test = SSSD.model(Y = Yi,R=R,Q=Q,
                  PAR=rad[i,],Xic=ICprior,THETA=theta[1:11],
                  dist = dist)
  save(test,dist,file=ofile)
  print(test$Analysis[[20]]$mu)
}
```

## visualization
```{r}
i = 181#181
ofile = paste0("OR_DP_",  ## DP = Disturbance Prior
                 formatC(i, width = 3, format = "d", flag = "0"),
                 "_P",Pi,"Q",Qi,"R",Ri,".RData")
load(ofile)
NT = floor(ncol(rad)/365)
Yi = matrix(NA,NT,3)  ## observed data
Yi[1:ncol(Yw),3] = Yw[i,]  ## only Cw observed
pool.names=c("Cleaf","Csoil","Cstem")
for(i in 1:3){

    #CI.Na = sapply(X = test$Analysis,function(x){
    #  y = x$CI
    #  sel.n = which(colnames(y) == paste0("n[",i,"]"))
    #  y[,sel.n]
    #})
    CI.Na = sapply(X = test$Analysis, function(x){
      quantile(x$Na[,i],c(0.025,0.25,0.5,0.75,0.975))
    })

@@

  plot(Yi[,i],
       ylim = range(c(range(c(range(CI.Na),Yi[,i]),na.rm=TRUE),0)),
       type='n',
       ylab=pool.names[i],xlab="time")
  ecoforecastR::ciEnvelope(1:ncol(CI.Nf),CI.Nf[1,],CI.Nf[3,],col=ecoforecastR::col.alpha("blue",0.5))
  ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[5,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
  lines(CI.Na[3,],col="#b2df8a",lwd=2,type='b',pch=18)
  points(Yi[,i])
  lines(CI.D[3,])
  if(ncol(CI.Na)==1){
    points(c(1,1),c(CI.Na[1,],CI.Na[5,]),pch="-",col="#b2df8a")
    points(c(1,1),c(CI.Nf[1,],CI.Nf[3,]),pch="-",col="blue")
    points(1,CI.D[3,],pch=17)
  }

@@

## process error
plot(test$bq, main="bq")
plot(test$aq, main="aq")
q = sapply(X = test$Analysis,function(x){
      y = x$mu["q"]
    })
plot(q)
plot(1/sqrt(q))
```
